---
title: "Lab 2"
author: "Jorge Mendez"
date: "2026-01-31"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
#install.packages("faraway")   # run once if you don't have it
library(faraway)

data(stat500)
head(stat500)

```

## Problem 2
### (a) Correlation Matrix

```{r}
round(cor(stat500), 2)

```
### (b)

```{r}
plot(total ~ final, data = stat500,
     pch = 16,
     xlab = "Final Exam Score",
     ylab = "Total Score",
     main = "Scatter Plot of Total vs Final")

```
### (c)


```{r}
model <- lm(total ~ final, data = stat500)
summary(model)

```
A simple linear regression model was fit with total score as the response variable and final exam score as the predictor. The estimated regression equation is

$\widehat{\text{total}} = 30.71 + 1.60(\text{final})$


The slope estimate indicates that for each one-point increase in final exam score, the total score increases by approximately 1.60 points on average. The slope is statistically significant (t = 9.04, p-value = 2.53 × 10⁻¹²).
