---
title: "Lab1"
author: "Jorge Mendez"
date: "2026-01-18"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Problems

## Problem 1:  Check Normality Dsitribution of data (QQ PLOT)

```{r, echo=FALSE}
sample <- c(26.4,23.5,25.4,22.9,25.2,39.2,25.5,31.9,26.0,44.6,35.5,38.6,
30.1,31.0,30.8,32.8,47.7,39.1,55.3,50.7,73.8,71.1,68.4,77.1,
19.4,19.3,18.7,19.0,23.2,21.3,23.2,19.9,18.9,19.8,19.6,21.9)

```

```{r, echo=FALSE}

hist(sample)
```

```{r}
sample.s <- sort(sample)  #sort data increasing
sample.s
```

```{r}
rank <- rank(sample.s) #rank data  from 1 to 36
rank
```
```{r}
size <- length(sample.s)   # tells me the size of the data
size
```

```{r}
p <- (rank-.5)/size
p
```
```{r}
z.quantile <- qnorm(p)     #Standard Normal Quantiles with such Probability
z.quantile 
```
#### QQ PLOT

```{r, echo=FALSE}
plot(x= z.quantile, y=sample.s, pch=16, main= 'QQ Plot')     #scatterplot of x=Z quantiles, y - data sorted
abline(lm(sample.s ~ z.quantile))
```


Question: Observe the patter of the points.

**Observation** : If the data were normal, deviations from the line would be small and random; here they are large, especially in the tails, indicating that the data is not approximately normal. 


```{r, echo=FALSE}
qqnorm(sample.s)
qqline(sample.s)
```


## Problem 2

```{r, echo=FALSE}
sample <-c(26.4,23.5,25.4,22.9,25.2,39.2,25.5,31.9,26.0,44.6,35.5,38.6,
30.1,31.0,30.8,32.8,47.7,39.1,55.3,50.7,73.8,71.1,68.4,77.1,
19.4,19.3,18.7,19.0,23.2,21.3,23.2,19.9,18.9,19.8,19.6,21.9)
```



```{r, echo=FALSE}
n <- length(sample)
xbar <- mean(sample)
s <- sd(sample)

n
xbar
s
```
a) Write a 95% CI for the population mean.

```{r, echo=FALSE}
alpha <- 0.05
tcrit <- qt(1 - alpha/2, df = n - 1)

ci_mean <- c(
  xbar - tcrit * s / sqrt(n),
  xbar + tcrit * s / sqrt(n)
)

ci_mean

```
**Interpretation:** We are 95% confident that the true population meanðœ‡lies within this interval.


b) Write a 95% CI for the population variance

```{r, echo=FALSE}
chisq_lower <- qchisq(alpha/2, df = n - 1)
chisq_upper <- qchisq(1 - alpha/2, df = n - 1)

ci_var <- c(
  (n - 1) * s^2 / chisq_upper,
  (n - 1) * s^2 / chisq_lower
)

ci_var

```
**Interpretation:** We are 95% confident that the true population variance lies between 187.8337 and 485.8378.


c) Write a 95% CI fot the population standard deviation


```{r, echo=FALSE}
ci_sd <- sqrt(ci_var)
ci_sd

```
**Assumption:** These confidence intervals are based on a random sample and (approximately) normal population assumptions; the variance and standard deviation intervals in particular rely on the normality assumption.


## Problem 3

```{r, echo=FALSE}

set.seed(123)

sample1 <- rnorm(50, mean=5, sd=2)
sample2 <- rnorm(70, mean=10, sd=5)

var.test(sample1, sample2)

```

- **Hâ‚€:** \(\sigma_1^2 = \sigma_2^2\)  
- **Hâ‚:** \(\sigma_1^2 \ne \sigma_2^2\)  
- **Conclusion:** Based on the p-value, we **reject Hâ‚€** at \(\alpha = 0.05\).



The F-test provides strong evidence against the null hypothesis of equal variances (p-value = 3.573eâˆ’09), indicating that the two population variances are not equal.


**Assumption:** This test assumes that both populations are normally distributed and that the samples are independent.



## Problem 4


```{r, echo=FALSE}
library(faraway)
data(stat500)

X <- stat500[, c("midterm", "final", "hw")]
head(X)
dim(X)
```

### Covariance matrix and correlation matrix

```{r, echo=FALSE}
cov_mat <- cov(X)
cor_mat <- cor(X)

cov_mat
cor_mat
round(cor_mat, 2)
```
**Interpretation:** The strongest linear relationship is between midterm and final scores (r = 0.55), while homework scores have weaker correlations with exam scores.

